---
title: "Differential Methylation Analysis"
author: "Sophia Sedille"
date: "2025-01-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries, include=FALSE}
# devtools::install_github('bmbolstad/preprocessCore', dependencies = T, upgrade = 'always', configure.args = '--disable-threading')

# Epigenetic packages
library(limma)
library(minfi)
library(EpiSmokEr)
library(IlluminaHumanMethylationEPICanno.ilm10b2.hg19)
library(IlluminaHumanMethylationEPICmanifest)
library(DMRcate)
library(missMethyl)
library(compareGroups)
library(ramwas)
library(FlowSorted.Blood.EPIC)
library(ChAMP)
library(openxlsx)
library(qvalue)

# Statistics
library(corpcor)

# Data handling and visualisation
library(openxlsx2)
library(tidyverse)
library(doParallel)
library(RColorBrewer)
library(Gviz)
library(gplots)
library(ggbeeswarm)
library(ggpmisc)
library(ggpubr)
library(see)
library(filematrix)
library(gprofiler2)
library(plotly)

recalculate <- FALSE
```


```{r user_paths}
# -------------------------------------------------------
# USER CONFIGURATION: Set these paths before running!
# -------------------------------------------------------

# Path to CSV file with PGS data (wide format, unscaled)
PATH_TO_PGS_DATA    <- "path/to/ADHD_PGSData_wide_unscaled.csv"

# Path to preprocessed targets RDS file
PATH_TO_TARGETS     <- "path/to/ProcessedData/targets.rds"

# Path to merged EPIC methylation data RDS file
PATH_TO_EPIC_MERGED <- "path/to/ProcessedData/all_EPIC_merged.rds"

# Output folder (must exist before running)
OUTPUT_FOLDER       <- "path/to/output/"

# -------------------------------------------------------
```


## Load and merge phenotypic data

```{r load_data}
pgs_data <- read_csv(PATH_TO_PGS_DATA)
pgs_data <- pgs_data[, c("Sample_Name", "ADHD")]  # keep only relevant columns

targets <- readRDS(PATH_TO_TARGETS)
targets$Sample_Name <- substr(targets$Sample_Name, 1, 8)  # remove T1 suffix

# Merge targets with PGS data
merged_table <- merge(targets, pgs_data, by = "Sample_Name")
colnames(merged_table)[colnames(merged_table) == "ADHD"] <- "ADHD_PGS"
merged_table$ADHD_PGS_scaled <- scale(merged_table$ADHD_PGS)
merged_table <- relocate(merged_table,
                         c(ADHD_PGS, ADHD_PGS_scaled, ADHD.average,
                           ADHD.hyperactivity, ADHD.inattention, SES_bin),
                         .after = Sample_ID)
```


## Preprocessing and calling of methylation data

```{r preprocessing}
allEPIC_merged <- readRDS(PATH_TO_EPIC_MERGED)

mqSet <- preprocessQuantile(allEPIC_merged)
mVals <- getM(mqSet)
bVals <- getBeta(mqSet)

rm(allEPIC_merged)
gc()
```


## Define SES variable

```{r define_ses}
targets$SES <- targets$Parental..highest..education.at.T1.in.education.categories.based.on.ISCED...based.on.school.leaving.qualification.and.training
targets$SES[targets$SES == 77] <- NA
targets$SES_bin <- factor(targets$SES > median(targets$SES, na.rm = TRUE),
                          levels = c(TRUE, FALSE),
                          labels = c("lowED", "highED"))
```


Some participants have multiple rows in `merged_table`. We check why:

```{r check_duplicates}
merged_table %>%
  filter(n() > 1, .by = Sample_Name) %>%
  View()
```

10 participants have 2 different PGS values for ADHD. We keep only the first entry per participant.

```{r deduplicate}
merged_table_dedup <- merged_table %>%
  slice_head(n = 1, by = Sample_Name)
```

Verify deduplication:

```{r check_dedup}
length(unique(merged_table_dedup$Sample_Name)) == nrow(merged_table_dedup)
```

```{r rownames}
rownames(merged_table_dedup) <- merged_table_dedup$index
```


## Differential methylation analysis using limma

```{r limma_analysis}
# Quick check: ADHD ~ covariates
lm(ADHD.average ~ Gender + Age + SES_bin, data = targets) %>% summary()

# Design matrix: triple interaction ADHD * SES * PGS
design <- model.matrix(
  ~1 + Gender + Age + Smoking_Scores + CD8T + CD4T + NK + Bcell + Mono + Neu +
    ADHD.average * SES_bin * ADHD_PGS_scaled,
  data = merged_table_dedup
)
mdata <- mVals[, rownames(design)]

# Fit linear model and apply empirical Bayes
fit  <- limma::lmFit(mdata, design)
fit2 <- limma::eBayes(fit)

# Extract results for the triple interaction term
res <- topTable(fit2, coef = "ADHD.average:SES_binhighED:ADHD_PGS_scaled", number = nrow(mdata))
head(res)

# Significant CpGs: p < 0.01 and |logFC| > 1.5
sig_cpg <- rownames(res)[res$P.Value < 0.01 & abs(res$logFC) > 1.5]
length(sig_cpg)

# Save decision test summary
summary_output <- capture.output(summary(decideTests(fit2, adjust.method = "none", p.value = 0.01)))
writeLines(summary_output, con = paste0(OUTPUT_FOLDER, "decideTests_summary_output.txt"))
```

```{r top_cpg}
# CpG with the smallest p-value among significant hits
sig_res      <- res[sig_cpg, ]
min_pval_cpg <- rownames(sig_res)[which.min(sig_res$P.Value)]
sig_res[which.min(sig_res$P.Value), ]
```


## Exploratory plots (not included in thesis)

```{r exploratory_plots}
sig_cpg_mvals <- t(mVals[sig_cpg, rownames(merged_table_dedup)])
tmp_data <- cbind(merged_table_dedup, sig_cpg_mvals)
tmp_data <- tmp_data %>% filter(!is.na(SES_bin))

tmp_data$ADHD_PGS_scaled_bin <- factor(
  tmp_data$ADHD_PGS_scaled > median(tmp_data$ADHD_PGS_scaled, na.rm = TRUE),
  levels = c(TRUE, FALSE), labels = c("low PGS", "high PGS")
)

plotlist <- list()
for (i in sig_cpg) {
  target_cpg <- i
  res_lm <- lm(as.formula(paste0(target_cpg, "~ 1 + Gender + Age + Smoking_Scores + CD8T + CD4T + NK + Bcell + Mono + Neu")),
               data = tmp_data)
  tmp_data[target_cpg] <- res_lm$residuals

  plotlist[[i]] <- ggplot(tmp_data, aes_string(x = "ADHD.average", y = target_cpg)) +
    geom_point() +
    facet_wrap(~SES_bin + ADHD_PGS_scaled_bin) +
    geom_smooth(method = "lm") +
    stat_fit_glance(method = "lm",
                    method.args = list(formula = y ~ x),
                    label.x = 0.5, label.y = 1,
                    aes(label = paste("p =", signif(..p.value.., 4),
                                      "  R² =", signif(..r.squared.., 3))),
                    size = 4)
}

combined_plot <- ggarrange(plotlist = plotlist, ncol = 2, nrow = 3)
ggsave(paste0(OUTPUT_FOLDER, "combined_plot.png"), plot = combined_plot[[1]], width = 12, height = 8, dpi = 300)
```


## Four CpGs of interest (used in thesis)

```{r four_cpgs}
cpgs_of_interest <- c("cg02700878", "cg17833746", "cg24312520", "cg14932016")
cpg_mvals <- t(mVals[cpgs_of_interest, rownames(merged_table_dedup)])
tmp_data  <- cbind(merged_table_dedup, cpg_mvals)

# Keep only complete cases for the relevant columns
key_cols <- c(cpgs_of_interest, "ADHD_PGS_scaled", "SES_bin", "ADHD.average")
tmp_data <- tmp_data[complete.cases(tmp_data[, key_cols]), ]

# Binary groupings
tmp_data$ADHD_bin <- factor(
  tmp_data$ADHD.average > median(tmp_data$ADHD.average, na.rm = TRUE),
  levels = c(TRUE, FALSE), labels = c("high ADHD", "low ADHD")
)
tmp_data$SES_ADHD_bin <- interaction(tmp_data$SES_bin, tmp_data$ADHD_bin, sep = " & ")
```


## Visualise association between ADHD-PGS and methylation across SES and ADHD severity groups

```{r visualize_cpgs}
plotlist <- list()
for (i in cpgs_of_interest) {
  target_cpg <- i

  # Regress out covariates and use residuals
  res_lm <- lm(as.formula(paste0(target_cpg, "~ 1 + Gender + Age + Smoking_Scores + CD8T + CD4T + NK + Bcell + Mono + Neu")),
               data = tmp_data)
  tmp_data[[target_cpg]] <- res_lm$residuals

  plotlist[[i]] <- ggplot(tmp_data, aes_string(x = "ADHD_PGS_scaled", y = target_cpg)) +
    geom_point(alpha = 0.6, size = 1.5) +
    facet_wrap(~SES_ADHD_bin) +
    geom_smooth(method = "lm", color = "slateblue", se = FALSE) +
    stat_fit_glance(method = "lm",
                    method.args = list(formula = y ~ x),
                    label.x = 0.5, label.y = 1,
                    aes(label = paste("p =", signif(..p.value.., 3),
                                      "| R² =", signif(..r.squared.., 3))),
                    size = 4) +
    labs(title = paste("CpG:", i), x = "ADHD PGS (scaled)", y = i) +
    theme_minimal(base_size = 12) +
    theme(strip.text = element_text(size = 12, face = "bold"),
          axis.title = element_text(size = 12),
          axis.text  = element_text(size = 10),
          plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
}

combined_plot <- ggarrange(plotlist[[1]], plotlist[[2]], plotlist[[3]], plotlist[[4]],
                           ncol = 2, nrow = 2)
ggsave(paste0(OUTPUT_FOLDER, "cpg_plots.pdf"), combined_plot, width = 12, height = 8)
```

```{r, fig.width=8, fig.height=12}
ggpubr::ggarrange(plotlist = plotlist, ncol = 2, nrow = 3)
```


## Annotate differentially methylated positions (DMPs)

```{r annotate_dmps}
anno_epic     <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b2.hg19)
anno_epic_sub <- anno_epic[match(rownames(mVals), anno_epic$Name),
                           c(1:4, 12:19, 24:ncol(anno_epic))]
rm(anno_epic)

DMPs <- topTable(fit2, num = Inf, coef = "ADHD.average:SES_binhighED:ADHD_PGS_scaled",
                 genelist = anno_epic_sub)
head(DMPs)

write_xlsx(DMPs, file = paste0(OUTPUT_FOLDER, "DMPs.xlsx"))
```


## Gene Set Enrichment Analysis of DMP genes using gprofiler2

```{r gsea_dmps}
gene_list <- unique(unlist(strsplit(DMPs$GencodeCompV12_NAME[DMPs$P.Value < 0.01], ";")))

gost_res <- gost(query = list("DMPs_average_SES_PGS" = gene_list), organism = "hsapiens")
head(gost_res$result)

p <- gostplot(gost_res, capped = FALSE, interactive = TRUE)
p

publish_gostplot(p, filename = paste0(OUTPUT_FOLDER, "gostplot_DMPs.pdf"))
```

Top 10 hits sorted by p-value:

```{r top10_hits}
go_results     <- gost_res$result
sorted_results <- go_results[order(go_results$p_value), ]
top10_results  <- sorted_results[1:10, c("term_name", "p_value", "intersection_size", "term_id")]
print(top10_results)

write.csv(go_results,    paste0(OUTPUT_FOLDER, "gostres_DMPs.csv"),       row.names = FALSE)
write.csv(top10_results, paste0(OUTPUT_FOLDER, "gostres_DMPs_top10.csv"), row.names = FALSE)
```


## Differentially Methylated Regions (DMRs) using DMRcate

```{r dmrcate_annotate}
my_annotation <- cpg.annotate(object = mdata, datatype = "array", what = "M",
                              annotation = c(array = "IlluminaHumanMethylationEPIC",
                                             annotation = "ilm10b2.hg19"),
                              analysis.type = "differential", design = design,
                              contrasts = FALSE,
                              coef = "ADHD.average:SES_binhighED:ADHD_PGS_scaled",
                              arraytype = "EPICv1")
```

```{r dmrcate_extract}
DMRs           <- dmrcate(my_annotation, lambda = 1000, C = 2, pcutoff = 0.01)
results_ranges <- extractRanges(DMRs)

# Extract chromosome names from coordinate string
chr_names <- sapply(strsplit(DMRs@coord, ":"), function(x) x[1])
results_ranges$seqnames <- chr_names

head(results_ranges$overlapping.genes)
head(results_ranges$seqnames)
```

```{r save_dmrs}
results_df <- as.data.frame(results_ranges)
write.csv(results_df, paste0(OUTPUT_FOLDER, "DMRs_results.csv"), row.names = FALSE)
```


## Gene Set Enrichment Analysis of DMR genes (FDR < 0.1)

```{r gsea_dmrs}
gene_list_dmr <- unique(unlist(strsplit(
  results_ranges$overlapping.genes[results_ranges$min_smoothed_fdr < 0.1], ","
)))

gost_res_dmr <- gost(query = gene_list_dmr, organism = "hsapiens")
head(gost_res_dmr$result)

p_dmr <- gostplot(gost_res_dmr, capped = FALSE, interactive = TRUE)
p_dmr
```

```{r save_gsea_dmrs}
# Convert list columns to character strings before saving
gost_clean <- gost_res_dmr$result
gost_clean[] <- lapply(gost_clean, function(col) {
  if (is.list(col)) sapply(col, toString) else col
})

write.csv(gost_clean, paste0(OUTPUT_FOLDER, "gostres_DMRs.csv"), row.names = FALSE)
```


## Convert DMRs to GRanges object

```{r granges}
dmr_gr <- GRanges(
  seqnames          = results_ranges$seqnames,
  ranges            = IRanges(start = start(results_ranges), end = end(results_ranges)),
  strand            = results_ranges$strand,
  min_smoothed_fdr  = results_ranges$min_smoothed_fdr,
  meandiff          = results_ranges$meandiff,
  overlapping_genes = results_ranges$overlapping.genes
)

options(showTailLines = Inf)
print(dmr_gr)
write.csv(as.data.frame(dmr_gr), paste0(OUTPUT_FOLDER, "DMRs_granges.csv"))
```
